Functional trait data tell new story about biodiversity change

### 2021-04-16 paper/manuscript.Rmd
Add corelation analysis, build discussion

We obtained mammal, bird, and amphibian timeseries from the Biotime
database, a global repository of high quality community assemblage
timeseries. All studies included in the database follow consistent
sampling protocols and represent full assemblages rather than population
of single species. [@dornelas2018]. Following best practices for the
database [@blowes2019] studies with multiple sample locations were split
into individual timeseries following a standardized scale. A global grid
was defined with a cell size determined based on the sample extent of
studies with only a single location (see [@dornelas2018] for details on
how sample extents were defined), with the area of each cell set to one
standard deviation away from the mean of the single extent locations.
All samples from a study within a single cell were considered to be a
single timeseries, and species abundances were combined for all samples.

@@

In order to ensure taxonomic consistency across datasets, Biotime
species were paired with trait data based on their species identifier
from the Integrated Taxonomic Information System database (citation),
obtained through the taxadb R package [@norman2020]. If more than one
species in the assemblage data resolved to the same identifier,
observations were considered the same species. For trait data, traits
for all species of the same identifier were averaged. Only studies with
at least 85% trait coverage were included. In order to have a sufficient
number of species to calculate functional diversity metrics years with
fewer than 4 species observed were excluded.

Many studies had a variable number of samples within years. To adjust
for this inconsistency in sampling effort we used sample-based
rarefaction by bootstrap resampling within years for each timeseries
based on the smallest number of samples in a year for that timeseries.

Our final dataset included 2439 timeseries from 54 studies in 21
countries and 15 biomes. Only four studies (consisting of 11 timeseries)
come from Amphibian studies due to the limited availability of amphibian
timeseries and low species richness values in those timeseries. For a
full breakdown of studies and their characteristics see the supplement.

@@

We corrected for bias in functional diversity metrics due to changes in
species richness by calculating the standardized effect size (SES) for
each metric from null estimates [@swenson2012]. Null model corrections
allow us to assess the degree to which the observed functional diversity
metric deviates from the value expected by chance in a randomly
assembled community. Null estimates were calculated for each rarefied
sample by randomly sampling species from the species pool for each year
and randomly assigning observed abundances to species. Species pools
included all species observed for a timeseries. This process was
repeated 500 times to get an estimate and standard deviation of the null
expectation for the metric for that timeseries. We used these values to
calculate SES using the following formula:
$SES = [F_{obs} - mean_{(F_{null})}]/SD_{(F_{null})}$. Unless otherwise
specified, functional diversity results are reported for standardized
values.

@@

@@

We assessed the type of changes happening in studies across metrics by
performing a cluster analysis of the study-level trends for each metric
using hierarchical clustering. Studies were clustered based on the
overall temporal slope for each metric. We included all standardized
functional diversity metrics and species richness to represent
independent facets of biodiversity change.
<!-- why do some studies not have blup estimates for FRic?-->

@@

We found no significant temporal trend in species richness or functional
diversity metrics for either observed or standardized metrics. The lack
of trend also held when data was broken down by taxa, biome and realm.
We did find a significant decrease in Jaccard similarity, indicating
increasing turnover through time.

At the study level, the majority of studies (78%) experienced a
significant temporal change in species richness and all functional
diversity metrics. Only four studies experienced no significant trend in
species richness clustering analysis showed that all but a few sites
were part of a cluster characterized by minimal deviance from a zero
temporal trend estimate for all metrics (Table 1). We identified four
other clusters containing six studies that were characterized by outlier
trends in at least one of the metrics and near-zero trends in the rest.
The first contained two studies with the most strongly decreasing
evenness. The second, two studies with the most strongly increasing
functional divergence and increasing functional richness. The third
contained one study with strongly decreasing functional divergence and
richness, and the fourth strongly decreasing species richness with
increasing functional richness.

@@

Our study represents the first broad-scale multi-taxa assessment of
functional change through time, giving a first look at overall trends in
functional diversity and the range of changes happening in mammal, bird,
and amphibian communities.

We found no systematic temporal change in any measured dimension of
functional diversity or species richness, even within specific realms,
biomes, or taxonomic groups. Species-based results are consistent with
other broad scale analyses of biodiversity change through time
[@dornelas2014] that found no overall trend in species richness despite
increasing turnover. While nearly all of the studies included in our
data experienced a significant increase or decrease in species richness,
they were observed in equal numbers resulting in an overall trend not
significantly different from zero. Surprisingly, functional diversity
metrics also did not capture any overall trends, indicating that similar
to species richness, communities are experience a wide range of
functional shifts that cannot be explained by their taxa, realm, or
climate.

>>>>

We obtained mammal, bird, and amphibian timeseries from the Biotime
database, a global repository of high quality community assemblage
timeseries. All studies included in the database follow consistent
sampling protocols and represent full assemblages rather than population
of single species. [@dornelas2018]. Following best practices for the
database [@blowes2019] studies with multiple sample locations were split
into individual timeseries following a standardized scale. Scale was set
by a global grid with cell size determined based on the sample extent of
studies with only a single location (see [@dornelas2018] for details on
how sample extents were defined), with the area of each cell set to one
standard deviation away from the mean of the single extent locations.
All samples from a study within a single cell were considered to be a
single timeseries, and species abundances were combined for all samples.

@@

In order to ensure taxonomic consistency across datasets, Biotime
species were paired with trait data based on their species identifier
from the Integrated Taxonomic Information System database (citation),
obtained through the taxadb R package [@norman2020]. If more than one
species in the assemblage data resolved to the same identifier,
observations were considered the same species. For trait data, traits
for all species of the same identifier were averaged. Only studies with
at least 75% trait coverage were included. In order to have a sufficient
number of species to calculate functional diversity metrics years with
fewer than 4 species observed were excluded.

Many studies had a variable number of samples within years. To account
for this inconsistency in sampling effort we used sample-based
rarefaction by bootstrap resampling within years for each timeseries
based on the smallest number of samples in a year for that timeseries.

Our final dataset included 2439 timeseries from 54 studies in 21
countries and 15 biomes. Only four studies (consisting of 11 timeseries)
come from Amphibian studies due to the limited availability of amphibian
timeseries and low species richness values for assemblages. For a full
breakdown of studies and their characteristics see the supplement.

@@

We corrected for bias in functional diversity metrics due to changes in
species richness by calculating the standardized effect size (SES) for
each metric from null estimates [@swenson2012]. Null model corrections
allow us to assess the degree to which the observed functional diversity
metric deviates from the value expected by chance in a randomly
assembled community. Null estimates were calculated for each rarefied
sample by randomly sampling species from the species pool for each year
and randomly assigning observed abundances to species. Species pools
included all species observed for a timeseries. This process was
repeated 500 times to get an estimate and standard deviation of the null
expectation for the metric for that timeseries. We used these values to
calculate SES using the following formula:
$SES = [F_{obs} - mean_{(F_{null})}]/SD_{(F_{null})}$. Unless otherwise
specified, functional diversity results are reported for standardized
values for the rest of the paper.

@@

*Overall Trends*

@@

*Study-level Trends*

We obtained study-level estimates of temporal change from the Best
Linear Unbiased Predictors (BLUP's) for each overall trend model and
assessed them in two ways. First, we looked at pairwise relationships
between all metrics to assess what kinds of community shifts are
commonly found together. We obtained correlations and bootstrapped
p-values using *corr.test* and *corCI* functions in the *psych* R
package (citation). Second, we assessed the relationship between all
metrics simultaneously by performing a cluster analysis which grouped
each study based on their similarities across all metrics. We used a
hierarchical clustering algorithm through the function *hclust* from the
R package *stats.* We included all standardized functional diversity
metrics, species richness, and Jaccard similarity in the analysis. We
report a qualitative interpretation of the groups from the cluster
analysis.

<!-- why do some studies not have blup estimates for FRic?-->

@@

We found no significant temporal trend in species richness or functional
diversity metrics for either observed or standardized metrics. There
continued to be no trend for studies broken down by taxa, biome or
realm. We did find a significant overall decrease in Jaccard similarity,
indicating increasing turnover through time.

At the study level, the majority of studies (78%) experienced a
significant temporal change in species richness and all functional
diversity metrics. 92% of sites experienced a significant trend for
species richness, 88% for functional functional richness, 96% for
functional divergence, 88% for functional evenness, and 35% for Jaccard
similarity. Only four studies experienced no significant trend in
species richness and were also characterized by significantly decreasing
functional evenness. There was only one significant pairwise correlation
between the metrics, a positive relationship between Functional
Divergence and Functional Evenness, indicating that studies experiencing
an increase in Divergence frequently experience and increase in
Evenness.

Cluster analysis grouped all but a few studies into a cluster
characterized by non-siginficant or very small significant slopes for
all metrics. We identified four other clusters containing six studies
that were characterized by outlier trends in at least one of the metrics
and near-zero trends in the rest. The first contained two studies with
the most strongly decreasing evenness. The second, two studies with the
most strongly increasing functional divergence and increasing functional
richness. The third contained one study with strongly decreasing
functional divergence and richness, and the fourth strongly decreasing
species richness with increasing functional richness.

@@

Our study represents the largest broad-scale multi-taxa assessment of
functional change through time, giving a first look at overall trends in
functional diversity and the range of changes happening in mammal, bird,
and amphibian communities. We found no systematic temporal change in any
measured dimension of functional diversity or species richness, even
within specific realms, biomes, or taxonomic groups. Species-based
results are consistent with other broad scale analyses of biodiversity
change through time [@dornelas2014] that found no overall trend in
species richness despite increasing turnover. While nearly all of the
studies included in our data experienced a significant increase or
decrease in species richness, they were observed in equal numbers
resulting in an overall trend not significantly different from zero.
Surprisingly, there was also no general trend in functional diversity
metrics indicating that, similar to species richness, communities are
experience a wide range of functional shifts. Non significant results
for study characteristics confirm that functional changes cannot be
explained by the taxa, realm, or climate characteristics of a given
study.

Individual studies in our dataset demonstrate the variety of species and
functional shifts that can happen concurrently in a community. Notably,
many studies did not experience increasing turnover (65%) despite having
significant change in functional dimensions.

-   very few sites experienced increasing turnover, but the ones that
    did were significant!

-   what does it me for turnover not to be increasing but the function
    of communities to be changing

-   Biological homogenization is one of the hypothesized meaningful
    forces
